|

### 2022-01-25 inst/MyArticle/MergePoster.Rmd
explicit call to funcitons with <package>::<fun>

bb <- data.frame(bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, y - yes but only for random effects, * - repeated measurements are only modeled using a fixed effects. ? - the hurdle model is not implemented in the msqrob2 package yet.")
```

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with _prolfqua_ in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and p-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `proflqua` (see Figure \@ref(fig:ContrastUML)).

@@

@@

Given a linear model contrasts can be computed by $\hat{\beta_{c}} = \sum l\beta_{m}$ and $\textrm{var} \hat\beta_c = l\sigma^2 (X^T X)^{-1} l^T$, with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of coefficients.  The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model.
For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `lmerTest` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

If there are no observation in a group we assume that there are no observations because of the limit of detection (LOD). Furthermore, we assume that the variance of the protein is constant in all the conditions.
We impute the mean estimate of the group by using the average of $10\%$ smallest group mean intensities in the experiment. To estimate the variance in the condition, we use the pooled variance based on all conditions:

$$s_p^2=\frac{\sum_{i=1}^k (n_i - 1)s_i^2}{\sum_{i=1}^k (n_i - 1)},$$
with $n_i$ the number of observations, and $s_i$ the standard deviation in each condition.
The standard deviation for the t-statics is then given by:

$$
s = \sqrt{2}\sqrt{\frac{n_g s_p^2}{n}},
$$

Where, $n_G$ is the number of groups and $n$ is the number of observations.

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom.  [@Smyth2004linear] discuss how, using the empirical Bayes paradigm, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior degrees of freedom, and posterior $\tilde \sigma$. These can be used to moderate the t-statistics by:
$\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}$
and subsequently the p-values. We implemented this method in the class `ContrastModerated` (Figure \@ref(fig:ContrastUML)).

@@

Afterwards, the median scaled p-value $\tilde{p}_s$ is determined and using the transformation below, transformed back onto the original scale.

@@

The following code example illustrates how we compute contrast comparing the protein intensities in the condition. First, the differences and the linear model are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we examine the importance of the factors in the model using the ANOVA function, which facilitates model selection for models with one explanatory variable. Finally, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply t-statistic moderation using the `ContrastModerated` class. Finally, the `addContrastResults` function merges both sets of contrast estimates, preferring the one obtained from the linear model. Figure \@ref(fig:exampleContrasts) Panel A shows the distribution of the p-values, while figure Panel B shows the volcano-plot for each comparison.

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the t-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p-value$ and moderated $p-value$. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), or false discovery proportion (FDP) which are given by:

@@

@@

Figure \@ref(fig:benchmarkROC) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t-statistics$ and scaled p-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the t-statistic or p-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

@@

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, for instance, to perform gene enrichment analysis, the t-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the p-values might improve the inference further. However, the observed change is minuscule.

@@

Computing the statistics at the peptide level, e.g., the t-statistics or p-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis. The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To adress the issue of missing observation we could either run proDA or the imputation of group averages method discussed here and afterwards apply ROPECA.

@@

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to (e.g., repeated measurements). If you use fixed-effect linear models, apply variance moderation since this improves the t-statistics and p-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the t-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

>>>>

bb <- data.frame(dplyr::bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, y - yes but only for random effects, * - repeated measurements are only modeled using a fixed effects. ? - the hurdle model is not implemented in the msqrob2 package yet.")
```

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with _prolfqua_ in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and p-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

@@

[TODO SAINTExpress integration, proDA]

@@

Given a linear model contrasts can be computed by $\hat{\beta_{c}} = \sum l\beta_{m}$ and $\textrm{var} \hat\beta_c = l\sigma^2 (X^T X)^{-1} l^T$, with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of coefficients.  The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model.

For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `lmerTest` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

If there are no observation in a group we assume that there are no observations because of the limit of detection (LOD). Furthermore, we assume that the variance of the protein is constant in all the conditions.
We impute the mean estimate of the group by using the average of $1\%$ smallest group mean intensities in the experiment. To estimate the variance in the condition, we use the pooled variance based on all conditions:

$$s_p^2=\frac{\sum_{i=1}^k (n_i - 1)s_i^2}{\sum_{i=1}^k (n_i - 1)},$$

with $n_i$ the number of observations, and $s_i$ the standard deviation in each condition.
The standard deviation for the $t$-statistics is then given by:

$$
s = \sqrt{\frac{2n_g s_p^2}{n}},
$$

Where, $n_g$ is the number of groups and $n$ is the number of observations.

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. [@Smyth2004linear] discuss how, using the empirical Bayes paradigm, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior $df$, and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:

$\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}$

and subsequently the $p$-values. We implemented this method in the class `ContrastModerated` (Figure \@ref(fig:ContrastUML)).

@@

Afterwards, the median scaled $p$-value $\tilde{p}_s$ is determined and using the transformation below, transformed back onto the original scale.

@@

The following code example illustrates how we compute contrast comparing the protein intensities in the condition. First, the differences and the linear model are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we examine the importance of the factors in the model using the ANOVA function, which facilitates model selection for models with one explanatory variable. Finally, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `addContrastResults` function merges both sets of contrast estimates, preferring the one obtained from the linear model. Figure \@ref(fig:exampleContrasts) Panel A shows the distribution of the $p$-values, while figure Panel B shows the volcano-plot for each comparison.

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p-value$ and moderated $p-value$. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), or false discovery proportion (FDP) which are given by:

@@

(TODO(CP): benchmark load from package - first build package with vignettes )
```{r}

@@

Figure \@ref(fig:benchmarkROC) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t-statistics$ and scaled p-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

@@

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, for instance, to perform gene enrichment analysis, the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference further. However, the observed change is minuscule.

@@

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis. The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To adress the issue of missing observation we could either run proDA or the imputation of group averages method discussed here and afterwards apply ROPECA.

@@

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to (e.g., repeated measurements). If you use fixed-effect linear models, apply variance moderation since this improves the t-statistics and p-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

### 2022-01-25 inst/MyArticle/PosterBioconductor.Rmd
explicit call to funcitons with <package>::<fun>

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with `prolfqua` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and p-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We also integrate the results of other packages, e.g. _SaintExpress_ or _proDA_. We use R6 to model the statistical modeling functionality in `proflqua` (see Figure \@ref(fig:ContrastUML)).

@@

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%'}
library(prolfqua)
d <- prolfqua_data('data_ionstar')$filtered()
lfqd <- LFQData$new(d$data, d$config) 
t <- lfqd$get_Transformer()
lfqd <-  t$log2()$robscale()$lfq
pl <- lfqd$get_Plotter()
p_1 <- pl$intensity_distribution_density() +
  labs(tag = "A") + theme(legend.position = "none")
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()
p_2 <- agr$plot()$plots[[1]] + labs(tag = "B")
p_3 <- lfqp$get_Stats()$violin() + labs(tag = "C")
pl <- lfqp$get_Plotter()
p_4 <- pl$boxplots()$boxplot[[1]] + labs(tag = "D")
ggpubr::ggarrange(p_1, p_2, p_3, p_4)
```

>>>>

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with `prolfqua` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and p-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We also integrate the results of other packages, e.g. _SaintExpress_ or _proDA_. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

@@

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%'}
library(prolfqua)
d <- prolfqua_data('data_ionstar')$filtered()
lfqd <- prolfqua::LFQData$new(d$data, d$config) 
t <- lfqd$get_Transformer()
lfqd <-  t$log2()$robscale()$lfq
pl <- lfqd$get_Plotter()
p_1 <- pl$intensity_distribution_density() +
  labs(tag = "A") + theme(legend.position = "none")
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()
p_2 <- agr$plot()$plots[[1]] + labs(tag = "B")
p_3 <- lfqp$get_Stats()$violin() + labs(tag = "C")
pl <- lfqp$get_Plotter()
p_4 <- pl$boxplots()$boxplot[[1]] + labs(tag = "D")
ggpubr::ggarrange(p_1, p_2, p_3, p_4)
```
